Signatures and Induction Principles for Higher Inductive-Inductive Types

### 2019-01-31 paper.tex
stuff

However, this is not the full picture. We would also like to have a
\emph{category} of algebras, with homomorphisms as morphisms, identity
morphisms, composition together with category laws. This is not without
difficulties.

We are working in an UIP-free setting. Here, standard definitions of categories
feature set-truncated morphisms \cite{HoTTbook}. But we have non-truncated
notions of algebras and morphisms, so we cannot use the standard
definitions. One solution is to simply use non-truncated categories; these have
been previously called ``precategories'' or ``wild categories''
\cite{semisegal}. Sojakova demonstrated \cite{sojakova} that working with wild
categories internally to type theory is enough to prove equivalence of induction
and homotopy initiality for a class of higher inductive types, which suggests
that the same might be possible for HIITs. Still, it would be desirable to build
semantics of HIITs in a richer ($\omega$, 1)-categorical setting.

However, out current approach does not scale to the point where we have to worry
about higher categories. We explain in the following. The logical next step
towards a categorical semantics would be defining a $\blank^\ID$ translation,
which computes identity homomorphisms:

@@

Here, the interpretation of terms witnesses functoriality: $t^\M$ maps
identity morphisms in $\Delta$ to displayed identity morphisms in $A$.
We translate $\U$ to identity functions:
\[
\U^\ID\,\hat{\gamma\,A} :\equiv \hat{\lambda\,(x : A).\, x}
\]
Note that $\U$ and $\U[\sigma]$ are definitionally equal in the theory of
signatures, where $\sigma$ is a parallel substitution. The translation of the
latter would be the following (omitting many details, including the handling of
substitutions in the translation):
\[
(\U[\sigma])^\ID\,\hat{\gamma\,A} \equiv \hat{\TR_{(\lambda\,x. A\ra A)}\,(\sigma^\ID\,\gamma)\,(\lambda\,x.\, x)}
\]
Here, $\hat{\sigma^\ID\,\gamma}$ yields an equation for functoriality of
$\hat{\sigma}$. Since the transport is constant, the result is propositionally
equal to $\hat{\lambda\,x.\, x}$. However, we need to preserve $\U = \U[\sigma]$
up to definitional equality.

@@

\section{Conclusions and further work}
\label{sec:summary}

>>>>

However, this is not the full picture. We would also like to have a
\emph{category} of algebras, with homomorphisms as morphisms. This is not
without difficulties.

We are working in an UIP-free setting. In such setting, standard definitions of
categories feature set-truncated morphisms \cite{HoTTbook}. But we have
non-truncated notions of algebras and homomorphisms, so we cannot use the
standard definitions. One solution is to simply use non-truncated categories;
these have been previously called ``precategories'' or ``wild categories''
\cite{semisegal}. Sojakova demonstrated \cite{sojakova} that working with wild
categories internally to type theory is enough to prove equivalence of induction
and homotopy initiality for a class of higher inductive types, which suggests
that the same might be possible for HIITs. Still, it would be desirable to build
semantics of HIITs in a richer ($\omega$, 1)-categorical setting.

However, out current approach does not scale to the point where we have to worry
about higher categories. We explain in the following. The natural next step
towards a categorical semantics would be defining a $\blank^\ID$ translation,
which computes identity homomorphisms:

@@

Here, the interpretation of terms witnesses functoriality: $t^\M$ maps
identity morphisms in $\Delta$ to displayed identity morphisms in $A$.
We translate $\U$ to identity functions:
\[
\U^\ID\,\hat{\gamma\,A} :\equiv \hat{\lambda\,(x : A).\, x}
\]
Note that $\U$ and $\U[\sigma]$ are definitionally equal in the theory of
signatures, where $\sigma$ is a parallel substitution. The translation of the
latter would be the following (omitting many details, including the handling of
substitutions in the translation):
\[
(\U[\sigma])^\ID\,\hat{\gamma\,A} \equiv \hat{\TR_{(\lambda\,x. A\ra A)}\,(\sigma^\ID\,\gamma)\,(\lambda\,x.\, x)}
\]
Here, $\hat{\sigma^\ID\,\gamma}$ yields an equation for functoriality of
$\hat{\sigma}$. Since the transport is constant, the result is propositionally
equal to $\hat{\lambda\,x.\, x}$. However, that is not enough, since we need to
preserve $\U = \U[\sigma]$ up to definitional equality.

There are numerous unavoidable instances of such failures of strict preservation
of definitional equalities for $\blank^\ID$, and they similarly arise when we
try to construct composition of homomorphisms. We consider three potential
solutions for this strictness problem:
\begin{enumerate}
\item Solving the coherence problem. This would allow us to interpret signatures
      into the metatheory, allowing preservation of definitional equality up
      to propositional metatheoretic equality.
\item Reformulating the syntax of signatures so that definitional equalities
      become weak propositional equalities. We already use weak $\beta$ for $\J$,
      can we do so elsewhere? However, this would result in an unusual and very
      inconvenient syntax of signatures, because we would need to weaken even
      basic substitution rules.
\item Instead of interpreting signatures into an UIP-free type theory, we
      interpret them into classical combinatorial structures for higher groupoids
      and categories, e.g.\ into simplicial sets. The drawback is that now we do not
      have the convenient synthetic notion of higher structures, provided by the syntax
      of type theory, and instead we have to manually build up these
      structures. Needless to say, this makes machine-checked formalization much
      more difficult. We have found mechanized formalization invaluable for the
      current paper, and it would be painful to abandon it in further research.
\end{enumerate}

In summary, solutions for the strictness problems require significant deviation
from the approach of the current paper, of require significant further research.

%% Nevertheless, the current approach of using syntactic translations is a sweet
%% spot for computing induction principles in a relatively simple way.

@@

\section{Conclusions}
\label{sec:summary}
